cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.Rmd
Updated trigometric functions

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles.

The corresponding $2\text{-}D$ coordinates are defined by:

$$
X_1 = \cos(\theta),
$$
$$
X_2 = \sin(\theta).
$$

This forms a nonlinear curvilinear structure in the shape of a tilted crescent in the $X_1$-$X_2$ plane.

For $p > 2$, additional dimensions $X_3$ to $X_p$ are generated using `gen_wavydims1()`:

$$
X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}.
$$

These noise dimensions are nonlinearly correlated with $\theta$, producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension:

Let $\theta \sim U(0, 3\pi)$ represent a random angle on a circular base and $z \sim U(0, h)$ represent the height along the cylinder.

The coordinates are defined as:

* $X_1 = \cos(\theta)$ (Circular base, x-axis),
* $X_2 = \sin(\theta)$ (Circular base, y-axis),
* $X_3 = z$ (Linear height),
* $X_4 = \sin(z)$ (Nonlinear curvy variation along height).

This forms a curvy cylindrical surface in $4\text{-}D$, where the fourth dimension bends periodically along the height axis, resembling a helicoid or twisting wave along the cylinder.

For $p > 4$, dimensions $X_5$ through $X_p$ are generated as:

$$
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \ldots, p.
$$

These dimensions represent unstructured, low-variance noise.

@@

Cartesian coordinates from spherical conversion:

  * $X_1 = \sin(\phi) \cdot \cos(\theta)$
  * $X_2 = \sin(\phi) \cdot \sin(\theta)$
  * $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim U(-0.5, 0.5)$ introduces vertical jitter.
  * $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path.

This generates a spherical spiral curve embedded in $4\text{-}D$ space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If $p > 4$, the function appends structured, non-linear noise via `gen_wavydims2()`. These dimensions are functions of the first coordinate $X_1$, introducing dependencies that preserve some geometric coherence. Each added dimension follows the form:

$$
X_j = s_j \cdot (-1)^{\lfloor j/2 \rfloor} \cdot X_1^{a_j} + \eta_j,
$$

where:
$a_j \in \{2, 3, 4, 5\}$ is a randomly chosen polynomial power,
$s_j \sim U(0.5, 2)$ is a scale factor,
$\eta_j \sim U(-\sigma_j, 2\sigma_j)$, with $\sigma_j \sim U(0, 0.05)$, adds mild randomness.

@@

The `gen_helicalspiral(n, p)` function generates a dataset of $n$ observations forming a helical spiral, embedded in the first four dimensions of a $p$-dimensional space. Additional dimensions (if $p > 4$) are filled with structured noise to mimic high-dimensional complexity.

The first four coordinates follow a $3\text{-}D$ helix with an additional oscillating component. Let $\theta \in [0, \frac{5\pi}{4}]$ be a sequence of angles controlling rotation around a circle.

Cartesian coordinates:
  * $X_1 = \cos(\theta)$: circular trajectory along the x-axis.

  * $X_2 = \sin(\theta)$: circular trajectory along the y-axis.

  * $X_3 = 0.05 \cdot \theta + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix.

  * $X_4 = 0.1 \cdot \sin(\theta)$: oscillates with $\theta$, representing a periodic "wobble" along the fourth dimension.

This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension.

When $p > 4$, the remaining $p - 4$ dimensions are populated using `gen_noisedims()`. These are independent Gaussian noise dimensions:

$$
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
$$

@@

The `gen_conicspiral(n, p, spins)` function generates a dataset of $n$ points forming a conical spiral in the first four dimensions of a $p$-dimensional space. The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a $3\text{-}D$ conic helix.

The shape is defined by parameter $\theta \in [0, 2\pi \cdot \text{spins}]$, controlling the angular progression of the spiral. The mapping to Cartesian space is as follows:

The Archimedean spiral in the horizontal plane is represented by;

$X_1 = \theta \cdot \cos(\theta)$ for radial expansion in x.
$X_2 = \theta \cdot \sin(\theta)$ for radial expansion in y.

The growth pattern resembles a cone, with the height increasing according to $X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.1, 0.6).$

Spiral modulation in the fourth dimension is represented by $X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim U(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension.

This results in a $3\text{-}D$ spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities.

For $p > 4$, the dataset includes isotropic Gaussian noise in the remaining dimensions via `gen_noisedims()`:

$$
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
$$

@@

The `gen_nonlinear(n, p, hc, non_fac)` function simulates a non-linear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables. 
The key geometry lies in dimensions 1–4, where:

$X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors).
$X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable.
$X_{2} = \frac{hc}{X_{1}} + \text{nonfac} \cdot \sin(X_{1})$: non-linear combination of **hyperbolic and sinusoidal transformations**, creating sharp curvature and oscillation.
$X_{4} = \cos(\pi \cdot X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure.

These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

If $p > 4$, the remaining dimensions are populated with Gaussian noise using `gen_noisedims()`:

$$
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
$$

>>>>

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles. The corresponding $2\text{-}D$ coordinates are defined by: $X_1 = \cos(\theta)$, $X_2 = \sin(\theta)$. For $p > 2$, additional dimensions $X_3$ to $X_p$ are generated using `gen_wavydims1()`: $X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}$. These noise dimensions are nonlinearly correlated with $\theta$, producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let $\theta \sim U(0, 3\pi)$ represent a random angle on a circular base and $z \sim U(0, h)$ represent the height along the cylinder. The coordinates are defined as: $X_1 = \cos(\theta)$ (Circular base, x-axis), $X_2 = \sin(\theta)$ (Circular base, y-axis), $X_3 = z$ (Linear height), and $X_4 = \sin(z)$ (Nonlinear curvy variation along height). This forms a curvy cylindrical surface in $4\text{-}D$, where the fourth dimension bends periodically along the height axis, resembling a helicoid or twisting wave along the cylinder. For $p > 4$, dimensions $X_5$ through $X_p$ are generated as: $X_j \sim N(0, 0.05^2),\text{ for }j = 5, \ldots, p$.

@@

Cartesian coordinates from spherical conversion: $X_1 = \sin(\phi) \cdot \cos(\theta)$, $X_2 = \sin(\phi) \cdot \sin(\theta)$, $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim U(-0.5, 0.5)$ introduces vertical jitter, and $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path. This generates a spherical spiral curve embedded in $4\text{-}D$ space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If $p > 4$, the function appends structured, non-linear noise via `gen_wavydims2()`. These dimensions are functions of the first coordinate $X_1$, introducing dependencies that preserve some geometric coherence. Each added dimension follows the form: $X_j = s_j \cdot (-1)^{\lfloor j/2 \rfloor} \cdot X_1^{a_j} + \eta_j$, where $a_j \in \{2, 3, 4, 5\}$ is a randomly chosen polynomial power, $s_j \sim U(0.5, 2)$ is a scale factor, $\eta_j \sim U(-\sigma_j, 2\sigma_j)$, with $\sigma_j \sim U(0, 0.05)$, adds mild randomness.

@@

The `gen_helicalspiral(n, p)` function generates a dataset of $n$ observations forming a helical spiral, embedded in the first four dimensions of $p\text{-}D$. Additional dimensions (if $p > 4$) are filled with structured noise to mimic high-dimensional complexity.

The first four coordinates follow a $3\text{-}D$ helix with an additional oscillating component. Let $\theta \in [0, \frac{5\pi}{4}]$ be a sequence of angles controlling rotation around a circle. Cartesian coordinates; $X_1 = \cos(\theta)$: circular trajectory along the x-axis, $X_2 = \sin(\theta)$: circular trajectory along the y-axis, $X_3 = 0.05 \cdot \theta + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix, and $X_4 = 0.1 \cdot \sin(\theta)$: oscillates with $\theta$, representing a periodic "wobble" along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension. When $p > 4$, the remaining $p - 4$ dimensions are populated using `gen_noisedims()`. These are independent Gaussian noise dimensions: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

@@

The `gen_conicspiral(n, p, spins)` function generates a dataset of $n$ points forming a conical spiral in the first four dimensions of $p\text{-}D$. The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a $3\text{-}D$ conic helix.

The shape is defined by parameter $\theta \in [0, 2\pi \cdot \text{spins}]$, controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; $X_1 = \theta \cdot \cos(\theta)$ for radial expansion in x, and $X_2 = \theta \cdot \sin(\theta)$ for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to $X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.1, 0.6).$ Spiral modulation in the fourth dimension is represented by $X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim U(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension. This results in a $3\text{-}D$ spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities. For $p > 4$, the dataset includes isotropic Gaussian noise in the remaining dimensions via `gen_noisedims()`: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

@@

The `gen_nonlinear(n, p, hc, non_fac)` function simulates a non-linear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables.

The $X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors), $X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable, $X_{2} = \frac{hc}{X_{1}} + \text{nonfac} \cdot \sin(X_{1})$: non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and $X_{4} = \cos(\pi \cdot X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter. If $p > 4$, the remaining dimensions are populated with Gaussian noise using `gen_noisedims()`: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

### 2025-06-24 paper-cardinalR.tex
Updated trigometric functions

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles.

The corresponding \(2\text{-}D\) coordinates are defined by:

\[
X_1 = \cos(\theta),
\]
\[
X_2 = \sin(\theta).
\]

This forms a nonlinear curvilinear structure in the shape of a tilted crescent in the \(X_1\)-\(X_2\) plane.

For \(p > 2\), additional dimensions \(X_3\) to \(X_p\) are generated using \texttt{gen\_wavydims1()}:

\[
X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}.
\]

These noise dimensions are nonlinearly correlated with \(\theta\), producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension:

Let \(\theta \sim U(0, 3\pi)\) represent a random angle on a circular base and \(z \sim U(0, h)\) represent the height along the cylinder.

The coordinates are defined as:

\begin{itemize}
\tightlist
\item
  \(X_1 = \cos(\theta)\) (Circular base, x-axis),
\item
  \(X_2 = \sin(\theta)\) (Circular base, y-axis),
\item
  \(X_3 = z\) (Linear height),
\item
  \(X_4 = \sin(z)\) (Nonlinear curvy variation along height).
\end{itemize}

This forms a curvy cylindrical surface in \(4\text{-}D\), where the fourth dimension bends periodically along the height axis, resembling a helicoid or twisting wave along the cylinder.

For \(p > 4\), dimensions \(X_5\) through \(X_p\) are generated as:

\[
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \ldots, p.
\]

These dimensions represent unstructured, low-variance noise.

@@

Cartesian coordinates from spherical conversion:

\begin{itemize}
\tightlist
\item
  \(X_1 = \sin(\phi) \cdot \cos(\theta)\)
\item
  \(X_2 = \sin(\phi) \cdot \sin(\theta)\)
\item
  \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim U(-0.5, 0.5)\) introduces vertical jitter.
\item
  \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path.
\end{itemize}

This generates a spherical spiral curve embedded in \(4\text{-}D\) space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If \(p > 4\), the function appends structured, non-linear noise via \texttt{gen\_wavydims2()}. These dimensions are functions of the first coordinate \(X_1\), introducing dependencies that preserve some geometric coherence. Each added dimension follows the form:

\[
X_j = s_j \cdot (-1)^{\lfloor j/2 \rfloor} \cdot X_1^{a_j} + \eta_j,
\]

where:
\(a_j \in \{2, 3, 4, 5\}\) is a randomly chosen polynomial power,
\(s_j \sim U(0.5, 2)\) is a scale factor,
\(\eta_j \sim U(-\sigma_j, 2\sigma_j)\), with \(\sigma_j \sim U(0, 0.05)\), adds mild randomness.

@@

The \texttt{gen\_helicalspiral(n,\ p)} function generates a dataset of \(n\) observations forming a helical spiral, embedded in the first four dimensions of a \(p\)-dimensional space. Additional dimensions (if \(p > 4\)) are filled with structured noise to mimic high-dimensional complexity.

The first four coordinates follow a \(3\text{-}D\) helix with an additional oscillating component. Let \(\theta \in [0, \frac{5\pi}{4}]\) be a sequence of angles controlling rotation around a circle.

Cartesian coordinates:
* \(X_1 = \cos(\theta)\): circular trajectory along the x-axis.

\begin{itemize}
\item
  \(X_2 = \sin(\theta)\): circular trajectory along the y-axis.
\item
  \(X_3 = 0.05 \cdot \theta + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix.
\item
  \(X_4 = 0.1 \cdot \sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension.
\end{itemize}

This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension.

When \(p > 4\), the remaining \(p - 4\) dimensions are populated using \texttt{gen\_noisedims()}. These are independent Gaussian noise dimensions:

\[
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
\]

@@

The \texttt{gen\_conicspiral(n,\ p,\ spins)} function generates a dataset of \(n\) points forming a conical spiral in the first four dimensions of a \(p\)-dimensional space. The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a \(3\text{-}D\) conic helix.

The shape is defined by parameter \(\theta \in [0, 2\pi \cdot \text{spins}]\), controlling the angular progression of the spiral. The mapping to Cartesian space is as follows:

The Archimedean spiral in the horizontal plane is represented by;

\(X_1 = \theta \cdot \cos(\theta)\) for radial expansion in x.
\(X_2 = \theta \cdot \sin(\theta)\) for radial expansion in y.

The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.1, 0.6).\)

Spiral modulation in the fourth dimension is represented by \(X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim U(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension.

This results in a \(3\text{-}D\) spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities.

For \(p > 4\), the dataset includes isotropic Gaussian noise in the remaining dimensions via \texttt{gen\_noisedims()}:

\[
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
\]

@@

The \texttt{gen\_nonlinear(n,\ p,\ hc,\ non\_fac)} function simulates a non-linear \(2\text{-}D\) surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables.
The key geometry lies in dimensions 1--4, where:

\(X_{1} \sim U(0.1, 2)\): base variable (avoids zero to prevent division errors).
\(X_{3} \sim U(0.1, 0.8)\): independent auxiliary variable.
\(X_{2} = \frac{hc}{X_{1}} + \text{nonfac} \cdot \sin(X_{1})\): non-linear combination of \textbf{hyperbolic and sinusoidal transformations}, creating sharp curvature and oscillation.
\(X_{4} = \cos(\pi \cdot X_{1}) + \varepsilon\), with \(\varepsilon \sim U(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure.

These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

If \(p > 4\), the remaining dimensions are populated with Gaussian noise using \texttt{gen\_noisedims()}:

\[
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
\]

>>>>

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles. The corresponding \(2\text{-}D\) coordinates are defined by: \(X_1 = \cos(\theta)\), \(X_2 = \sin(\theta)\). For \(p > 2\), additional dimensions \(X_3\) to \(X_p\) are generated using \texttt{gen\_wavydims1()}: \(X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}\). These noise dimensions are nonlinearly correlated with \(\theta\), producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let \(\theta \sim U(0, 3\pi)\) represent a random angle on a circular base and \(z \sim U(0, h)\) represent the height along the cylinder. The coordinates are defined as: \(X_1 = \cos(\theta)\) (Circular base, x-axis), \(X_2 = \sin(\theta)\) (Circular base, y-axis), \(X_3 = z\) (Linear height), and \(X_4 = \sin(z)\) (Nonlinear curvy variation along height). This forms a curvy cylindrical surface in \(4\text{-}D\), where the fourth dimension bends periodically along the height axis, resembling a helicoid or twisting wave along the cylinder. For \(p > 4\), dimensions \(X_5\) through \(X_p\) are generated as: \(X_j \sim N(0, 0.05^2),\text{ for }j = 5, \ldots, p\).

@@

Cartesian coordinates from spherical conversion: \(X_1 = \sin(\phi) \cdot \cos(\theta)\), \(X_2 = \sin(\phi) \cdot \sin(\theta)\), \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim U(-0.5, 0.5)\) introduces vertical jitter, and \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path. This generates a spherical spiral curve embedded in \(4\text{-}D\) space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If \(p > 4\), the function appends structured, non-linear noise via \texttt{gen\_wavydims2()}. These dimensions are functions of the first coordinate \(X_1\), introducing dependencies that preserve some geometric coherence. Each added dimension follows the form: \(X_j = s_j \cdot (-1)^{\lfloor j/2 \rfloor} \cdot X_1^{a_j} + \eta_j\), where \(a_j \in \{2, 3, 4, 5\}\) is a randomly chosen polynomial power, \(s_j \sim U(0.5, 2)\) is a scale factor, \(\eta_j \sim U(-\sigma_j, 2\sigma_j)\), with \(\sigma_j \sim U(0, 0.05)\), adds mild randomness.

@@

The \texttt{gen\_helicalspiral(n,\ p)} function generates a dataset of \(n\) observations forming a helical spiral, embedded in the first four dimensions of \(p\text{-}D\). Additional dimensions (if \(p > 4\)) are filled with structured noise to mimic high-dimensional complexity.

The first four coordinates follow a \(3\text{-}D\) helix with an additional oscillating component. Let \(\theta \in [0, \frac{5\pi}{4}]\) be a sequence of angles controlling rotation around a circle. Cartesian coordinates; \(X_1 = \cos(\theta)\): circular trajectory along the x-axis, \(X_2 = \sin(\theta)\): circular trajectory along the y-axis, \(X_3 = 0.05 \cdot \theta + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix, and \(X_4 = 0.1 \cdot \sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension. When \(p > 4\), the remaining \(p - 4\) dimensions are populated using \texttt{gen\_noisedims()}. These are independent Gaussian noise dimensions: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).

@@

The \texttt{gen\_conicspiral(n,\ p,\ spins)} function generates a dataset of \(n\) points forming a conical spiral in the first four dimensions of \(p\text{-}D\). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a \(3\text{-}D\) conic helix.

The shape is defined by parameter \(\theta \in [0, 2\pi \cdot \text{spins}]\), controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; \(X_1 = \theta \cdot \cos(\theta)\) for radial expansion in x, and \(X_2 = \theta \cdot \sin(\theta)\) for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.1, 0.6).\) Spiral modulation in the fourth dimension is represented by \(X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim U(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension. This results in a \(3\text{-}D\) spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities. For \(p > 4\), the dataset includes isotropic Gaussian noise in the remaining dimensions via \texttt{gen\_noisedims()}: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).

@@

The \texttt{gen\_nonlinear(n,\ p,\ hc,\ non\_fac)} function simulates a non-linear \(2\text{-}D\) surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables.

The \(X_{1} \sim U(0.1, 2)\): base variable (avoids zero to prevent division errors), \(X_{3} \sim U(0.1, 0.8)\): independent auxiliary variable, \(X_{2} = \frac{hc}{X_{1}} + \text{nonfac} \cdot \sin(X_{1})\): non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and \(X_{4} = \cos(\pi \cdot X_{1}) + \varepsilon\), with \(\varepsilon \sim U(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter. If \(p > 4\), the remaining dimensions are populated with Gaussian noise using \texttt{gen\_noisedims()}: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).
